
# Data preparation

Load the LEVANTE survey data and define site labels used throughout.

```{r}
# https://github.com/levante-framework/levante-r
library(levante)
library(psych)      
library(patchwork)
library(tidyverse)
library(ggthemes)

surveys <- get_surveys(data_source = "levante-data-pilots-hackathon",
                       version = "current")

# sites present in the data, used in plots and subsetting
sites    <- surveys |> distinct(team) |> pull(team)
site_pal <- solarized_pal()(length(sites)) |> set_names(sites)
```

# Caregiver survey analysis

## Data preparation

Filter to caregiver measurement items and deduplicate: keep each caregiver's
earliest survey session, then take one response per item within that session.

```{r}
#| label: caregiver-setup
caregiver_raw <- surveys |>
  filter(survey_type == "caregiver",
         question_type == "Measurement",
         !is.na(form_construct),
         !is.na(value)) |>
  mutate(site = team) |>
  filter(site %in% sites) |>
  group_by(site, child_id) |>
  filter(survey_timestamp == min(survey_timestamp)) |>   # earliest session per caregiver
  ungroup() |>
  group_by(site, child_id, form_construct, variable) |>
  slice(1) |>   # one response per item (handles exact timestamp ties)
  ungroup()
```

## Parenting

The Parenting scale covers three subconstructs: **Warmth** (5 items),
**Control** (5 items), and **Stress** (4 items), all on a 5-point scale (0–4).
We use EFA to examine whether the theoretical factor structure is recovered
empirically across sites.

### Data selection

Extract item-to-subconstruct metadata, then reshape to wide format keeping
only complete cases.

```{r}
#| label: pt-setup
# item metadata: variable → subconstruct mapping
pt_meta <- caregiver_raw |>
  filter(form_construct == "Parenting") |>
  distinct(variable, subconstruct = form_subconstruct)

pt_items      <- pt_meta |> pull(variable)
item_order_pt <- pt_meta |> arrange(subconstruct) |> pull(variable)

pt <- caregiver_raw |>
  filter(form_construct == "Parenting") |>
  select(site, child_id, variable, value) |>
  pivot_wider(names_from = variable, values_from = value) |>
  filter(complete.cases(across(all_of(pt_items))))   # factanal() requires no missing values
```

### Response option distribution

With five response categories, these Likert-type items are conventionally
treated as continuous (interval-level) rather than ordinal in the factor
analysis below. This approximation is generally considered acceptable once a
scale has five or more categories and responses are not severely skewed
(Rhemtulla et al., 2012), which motivates the maximum-likelihood approach
(`factanal()`) used here rather than a polychoric-based alternative.

```{r}
#| label: pt-response-dist
#| fig-width: 11
#| fig-height: 5
pt |>
  pivot_longer(all_of(pt_items), names_to = "variable", values_to = "value") |>
  count(site, variable, value) |>
  group_by(site, variable) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  mutate(variable = factor(variable, levels = rev(item_order_pt))) |>
  ggplot(aes(x = prop, y = variable, fill = factor(value))) +
  geom_col() +
  facet_wrap(~ site, nrow = 1) +
  scale_fill_viridis_d(name = "Response", option = "mako", direction = -1) +
  labs(x = "Proportion", y = NULL) +
  theme_classic() +
  theme(legend.position = "bottom")
```

### Standardization

Z-score all items pooled across sites so loadings are on a common scale.


```{r}
#| label: pt-scale
pt_scaled <- pt |>
  mutate(across(all_of(pt_items), \(x) as.numeric(scale(x))))
```

### Parallel analysis

Run parallel analysis separately for each site. Compare observed factor
eigenvalues against eigenvalues from random data of the same size — factors
above the simulated line are retained.

```{r}
#| label: pt-parallel
par_results_pt <- list()
scree_data_pt  <- list()

for (sitei in sites) {
  pt_site <- pt_scaled |>
    filter(site == sitei) |>
    select(all_of(pt_items))

  res <- fa.parallel(pt_site, fa = "fa", n.iter = 100, plot = FALSE)
  par_results_pt[[sitei]] <- res

  ncomp <- length(res$fa.values)
  scree_data_pt[[sitei]] <- tibble(
    site       = sitei,
    factor     = rep(1:ncomp, 2),
    eigenvalue = c(res$fa.values, res$fa.sim),
    type       = rep(c("Observed", "Simulated (mean)"), each = ncomp)
  )
}

# compare with the scree plot below to confirm
nfactors_pt <- sapply(par_results_pt, \(r) r$nfact)
print(nfactors_pt)
```

The scree plot below shows observed eigenvalues (solid blue) versus simulated
eigenvalues from random data (dashed red). The parallel analysis criterion retains
factors where the observed line is above the simulated line. Use this plot
together with the algorithmic suggestion above to decide on the final factor count.

```{r}
#| label: pt-scree
#| fig-width: 10
#| fig-height: 4
bind_rows(scree_data_pt) |>
  ggplot(aes(x = factor, y = eigenvalue, color = type, linetype = type)) +
  geom_line() +
  geom_point(aes(shape = type)) +
  facet_wrap(~ site, nrow = 1) +
  scale_color_manual(values = c("Observed" = "#268bd2", "Simulated (mean)" = "#dc322f")) +
  scale_linetype_manual(values = c("Observed" = "solid", "Simulated (mean)" = "dashed")) +
  scale_shape_manual(values = c("Observed" = 16, "Simulated (mean)" = 1)) +
  labs(x = "Factor", y = "Eigenvalue", color = NULL, linetype = NULL, shape = NULL) +
  theme_classic() +
  theme(legend.position = "bottom")
```

Parallel analysis consistently suggests 3 factors across all three sites,
consistent with the scale's 3 theoretical subconstructs (Warmth, Control, Stress).
The scree elbow at factor 3 further corroborates this choice.

### Exploratory factor analysis

We fit EFA with **promax rotation** (oblique — factors are allowed to correlate).
Promax is preferred over varimax (orthogonal) here because Warmth, Control, and
Stress are theoretically related and forcing them to be uncorrelated would be
unrealistic.


Only the highest loading per item is displayed; others are zeroed out.
This works cleanly here because all reverse-scored items in levante survey data 
have already been recoded so all loadings are expected to be positive.

```{r}
#| label: pt-efa
#| fig-width: 12
#| fig-height: 6
plot_list_pt <- list()

for (sitei in sites) {
  pt_site <- pt_scaled |>
    filter(site == sitei) |>
    select(all_of(pt_items))

  if (nrow(pt_site) < 20) next

  efa_res <- factanal(
    ~ ., data = pt_site,
    factors  = nfactors_pt[sitei],
    rotation = "promax",
    na.action = na.exclude
  )

  # unclass() strips the special "loadings" class so it can be coerced to a data frame
  load_df <- as.data.frame(unclass(efa_res$loadings)) |>
    rownames_to_column("variable") |>
    pivot_longer(-variable, names_to = "factor", values_to = "loading") |>
    mutate(factor = gsub("Factor", "F", factor)) |>
    group_by(variable) |>
    mutate(loading = if_else(loading == max(loading), loading, 0)) |>
    ungroup() |>
    left_join(pt_meta, by = "variable") |>
    mutate(variable = factor(variable, levels = rev(item_order_pt)))

  sub_df  <- distinct(load_df, variable, subconstruct)
  is_last <- sitei == tail(sites, 1)

  p <- ggplot(load_df, aes(x = factor, y = variable, fill = loading)) +
    geom_tile(color = "white") +
    geom_text(aes(label = if_else(loading != 0, sprintf("%.2f", loading), "")),
              size = 3) +
    scale_fill_gradient2(low = "#2c7bb6", high = "#d7191c", mid = "white",
                         midpoint = 0, name = "Loading") +
    theme_classic() +
    labs(x = NULL, y = "Item", title = sitei) +
    theme(legend.position = "none")

  if (is_last) {
    # add subconstruct labels on the right of the last panel only
    p <- p +
      geom_text(data = sub_df,
                aes(x = Inf, y = variable, label = subconstruct),
                inherit.aes = FALSE, hjust = -0.05, size = 3) +
      coord_cartesian(clip = "off") +
      theme(plot.margin = margin(r = 120))
  } else if (sitei != sites[1]) {
    p <- p +
      coord_cartesian(clip = "off") +
      theme(axis.text.y = element_blank(), axis.title.y = element_blank(),
            plot.margin = margin(r = 10))
  } else {
    p <- p + coord_cartesian(clip = "off") + theme(plot.margin = margin(r = 10))
  }

  plot_list_pt[[sitei]] <- p
}

wrap_plots(plot_list_pt, nrow = 1) +
  plot_annotation(
    title = "Parenting EFA",
    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
  )
```

# Child survey analysis

EFA is an exploratory tool: it tells us whether a factor structure exists in
the data. Once we have a theoretically motivated structure — either from EFA
results or prior literature — we can use **Confirmatory Factor Analysis
(CFA)**, which tests whether a pre-specified structure fits the data.

When data come from multiple sites or groups, CFA also lets us ask a further
question: does the scale measure the same construct in the same way across
groups? This is **measurement invariance testing**, and it is the focus of
the rest of this tutorial.

We demonstrate the full pipeline on two child survey constructs: **Belonging**,
**Peer Relationship**.

```{r}
#| label: child-setup
library(lavaan)

# load all child measurement items once; filter per construct below
child_raw <- surveys |>
  filter(survey_type == "child",
         question_type == "Measurement",
         !is.na(value)) |>
  mutate(site = team) |>
  filter(site %in% sites) |>
  group_by(site, child_id) |>
  filter(survey_timestamp == min(survey_timestamp)) |>
  ungroup() |>
  group_by(site, child_id, variable) |>
  slice(1) |>
  ungroup()
```

For all three constructs we follow the same sequence:

1. Check response distributions — WLSMV requires all response categories to
   appear in every site (empty cells prevent polychoric correlation estimation).
2. Fit the **configural** model — all parameters free per site.
3. Fit the **metric** model — loadings constrained equal; compute ΔCFI vs. configural.
4. Fit the **scalar** model — thresholds also constrained equal; compute ΔCFI vs. metric.
5. Visualise final parameter estimates as a heatmap.

The identification strategy throughout: fix factor variance to 1 (`F ~~ 1*F`)
and free all loadings (`NA*` on the first item).

---

## Belonging

Belonging (3 items, 4-point scale 0–3) asks children how connected and happy
they feel at school.

Filter and reshape the data to wide format with one row per child.

```{r}
#| label: bl-prep
bl <- child_raw |>
  filter(form_construct == "Belonging") |>
  select(site, child_id, variable, value) |>
  pivot_wider(names_from = variable, values_from = value)

bl_items <- setdiff(names(bl), c("site", "child_id"))
bl <- bl |> filter(complete.cases(across(all_of(bl_items))))
```

WLSMV requires every response category to appear in every site. Check that no category is completely absent for any item.

```{r}
#| label: bl-response-dist
#| fig-width: 9
#| fig-height: 3
child_raw |>
  filter(form_construct == "Belonging") |>
  count(site, variable, value) |>
  group_by(site, variable) |>
  mutate(percent = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = factor(value), y = percent, fill = site)) +
  geom_col(position = "dodge") +
  facet_wrap(~ variable, nrow = 1) +
  scale_fill_manual(values = site_pal) +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(x = "Response", y = "Proportion", title = "Belonging") +
  theme_classic() + theme(legend.position = "bottom")
```

Specify the one-factor model in lavaan syntax. `NA*` on the first item tells lavaan to estimate that loading freely. `BL ~~ 1*BL` fixes the factor variance to 1, which is required to give the model a scale.

```{r}
#| label: bl-model
model_bl <- '
  BL =~ NA*SchoolEnjoy + SchoolFun + SchoolHappy
  BL ~~ 1*BL
'
```

Fit the **configural model** — the same factor structure in all sites, but every loading and threshold is estimated separately per site. This is the unconstrained baseline; all subsequent models are compared against it.

```{r}
#| label: bl-configural
#| warning: false
fit_configural_bl <- cfa(model_bl, data = bl, group = "site",
                         estimator = "WLSMV", ordered = bl_items)
fitMeasures(fit_configural_bl, c("cfi", "tli", "rmsea", "srmr"))
```

Add loading equality constraints across sites (**metric model**). A positive ΔCFI > .01 would mean the items do not relate to the factor with equal strength in every site.

```{r}
#| label: bl-metric
#| warning: false
fit_metric_bl <- cfa(model_bl, data = bl, group = "site",
                     group.equal = "loadings",
                     estimator = "WLSMV", ordered = bl_items)
fitMeasures(fit_metric_bl, c("cfi", "tli", "rmsea", "srmr"))

# positive = CFI dropped (worse fit)
fitMeasures(fit_configural_bl, "cfi") - fitMeasures(fit_metric_bl, "cfi")
```

Add threshold equality constraints on top of the metric model (**scalar model**). A positive ΔCFI > .01 would mean children in different sites use the response scale differently for the same level of the latent trait.

```{r}
#| label: bl-scalar
#| warning: false
fit_scalar_bl <- cfa(model_bl, data = bl, group = "site",
                     group.equal = c("loadings", "thresholds"),
                     estimator = "WLSMV", ordered = bl_items)
fitMeasures(fit_scalar_bl, c("cfi", "tli", "rmsea", "srmr"))

fitMeasures(fit_metric_bl, "cfi") - fitMeasures(fit_scalar_bl, "cfi")
```

Full metric and scalar invariance hold (ΔCFI ≤ .01 at both steps).
`fit_scalar_bl` is the final model.

Print the full model summary — parameter estimates for all loadings and thresholds across sites.

```{r}
#| label: bl-summary
summary(fit_scalar_bl)
```

Visualise loadings (λ) and thresholds (τ) as a heatmap. Each column is a parameter type; each row is an item. When the scale is fully invariant, values in each row should be nearly identical across sites.

```{r}
#| label: bl-params
#| fig-width: 8
#| fig-height: 3
pe_bl <- parameterEstimates(fit_scalar_bl)
gl_bl <- lavInspect(fit_scalar_bl, "group.label")

params_bl <- bind_rows(
  pe_bl |> filter(op == "=~") |>
    transmute(site  = factor(group, labels = gl_bl),
              item  = rhs, value = est, param = "lambda"),
  pe_bl |> filter(op == "|") |>
    transmute(site  = factor(group, labels = gl_bl),
              item  = lhs, value = est,
              param = paste0("tau", gsub("t", "", rhs)))
)

ggplot(params_bl, aes(param, item, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", value)), size = 3.5) +
  scale_fill_gradient2(low = "#2c7bb6", mid = "white",
                       high = "#d7191c", midpoint = 0, name = "Estimate") +
  scale_x_discrete(limits = c("lambda", "tau1", "tau2", "tau3"),
                   labels = c(lambda = expression(lambda),
                              tau1   = expression(tau[1]),
                              tau2   = expression(tau[2]),
                              tau3   = expression(tau[3]))) +
  facet_wrap(~ site, nrow = 1) +
  theme_classic() +
  labs(x = NULL, y = "Item")
```

---

## Peer Relationship

Peer Relationship captures the quality of children's friendships and social
interactions with classmates.

Filter and reshape the Peer Relationships data to wide format with one row per child.

```{r}
#| label: pr-prep
pr <- child_raw |>
  filter(form_construct == "Peer Relationships") |>
  select(site, child_id, variable, value) |>
  pivot_wider(names_from = variable, values_from = value)

pr_items <- setdiff(names(pr), c("site", "child_id"))
pr <- pr |> filter(complete.cases(across(all_of(pr_items))))
```

Check the response distribution for each item across sites. Look for floor or ceiling effects before modelling.

```{r}
#| label: pr-response-dist
#| fig-width: 9
#| fig-height: 3
child_raw |>
  filter(form_construct == "Peer Relationships") |>
  count(site, variable, value) |>
  group_by(site, variable) |>
  mutate(percent = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = factor(value), y = percent, fill = site)) +
  geom_col(position = "dodge") +
  facet_wrap(~ variable, nrow = 1) +
  scale_fill_manual(values = site_pal) +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(x = "Response", y = "Proportion", title = "Peer Relationship") +
  theme_classic() + theme(legend.position = "bottom")
```

Define the one-factor CFA model. Item names come directly from the data — no hardcoding needed.

```{r}
#| label: pr-model
model_pr <- paste0(
  "PR =~ NA*", pr_items[1], " + ", paste(pr_items[-1], collapse = " + "), "\n",
  "PR ~~ 1*PR"
)
cat(model_pr)
```

Fit the **configural model**: the same factor structure is imposed in each site, but all parameters are estimated freely per site.

```{r}
#| label: pr-configural
#| warning: false
fit_configural_pr <- cfa(model_pr, data = pr, group = "site",
                         estimator = "WLSMV", ordered = pr_items)
fitMeasures(fit_configural_pr, c("cfi", "tli", "rmsea", "srmr"))
```

Add loading equality constraints across sites (**metric model**). A positive ΔCFI > .01 indicates non-invariant loadings.

```{r}
#| label: pr-metric
#| warning: false
fit_metric_pr <- cfa(model_pr, data = pr, group = "site",
                     group.equal = "loadings",
                     estimator = "WLSMV", ordered = pr_items)
fitMeasures(fit_metric_pr, c("cfi", "tli", "rmsea", "srmr"))

# positive = CFI dropped
fitMeasures(fit_configural_pr, "cfi") - fitMeasures(fit_metric_pr, "cfi")
```

The score test ranks loadings by how much they worsen fit when constrained. The item in the top row is the one to free.

```{r}
#| label: pr-metric-score
#| warning: false
pt_pr <- parTable(fit_metric_pr) |> filter(op == "=~")

lavTestScore(fit_metric_pr)$uni |>
  mutate(item = pt_pr$rhs[match(lhs, pt_pr$label)]) |>
  filter(!is.na(item)) |>
  arrange(desc(X2))
```

Free the most non-invariant loading and refit (**partial metric model**).

```{r}
#| label: pr-metric-partial
#| warning: false
fit_metric_partial_pr <- cfa(model_pr, data = pr, group = "site",
                             group.equal = "loadings",
                             group.partial = "PR=~LonelySchool",
                             estimator = "WLSMV", ordered = pr_items)
fitMeasures(fit_metric_partial_pr, c("cfi", "tli", "rmsea", "srmr"))

fitMeasures(fit_configural_pr, "cfi") - fitMeasures(fit_metric_partial_pr, "cfi")
```

Add threshold equality constraints on top of the partial metric model (**scalar model**). ΔCFI > .01 means thresholds are also non-invariant.

```{r}
#| label: pr-scalar
#| warning: false
fit_scalar_pr <- cfa(model_pr, data = pr, group = "site",
                     group.equal = c("loadings", "thresholds"),
                     group.partial = "PR=~LonelySchool",
                     estimator = "WLSMV", ordered = pr_items)
fitMeasures(fit_scalar_pr, c("cfi", "tli", "rmsea", "srmr"))

fitMeasures(fit_metric_partial_pr, "cfi") - fitMeasures(fit_scalar_pr, "cfi")
```

Score test on the scalar model to identify the most non-invariant threshold.

```{r}
#| label: pr-scalar-score
#| warning: false
pt_sc_pr <- parTable(fit_scalar_pr) |>
  filter(op == "|", group == 1) |>
  mutate(gp_str = paste0(lhs, "|", rhs))

lavTestScore(fit_scalar_pr)$uni |>
  mutate(gp_str = pt_sc_pr$gp_str[match(lhs, pt_sc_pr$label)]) |>
  filter(!is.na(gp_str)) |>
  arrange(desc(X2))
```

Free the top non-invariant threshold (first partial scalar correction).

```{r}
#| label: pr-scalar-partial
#| warning: false
fit_scalar_partial_pr <- cfa(model_pr, data = pr, group = "site",
                             group.equal = c("loadings", "thresholds"),
                             group.partial = c("PR=~LonelySchool", "ClassNice | t3"),
                             estimator = "WLSMV", ordered = pr_items)
fitMeasures(fit_scalar_partial_pr, c("cfi", "tli", "rmsea", "srmr"))

fitMeasures(fit_metric_partial_pr, "cfi") - fitMeasures(fit_scalar_partial_pr, "cfi")
```

ΔCFI is still > .01 — run the score test again on the partially corrected model to find the next non-invariant threshold.

```{r}
#| label: pr-scalar-score-2
#| warning: false
pt_sc2_pr <- parTable(fit_scalar_partial_pr) |>
  filter(op == "|", group == 1) |>
  mutate(gp_str = paste0(lhs, "|", rhs))

lavTestScore(fit_scalar_partial_pr)$uni |>
  mutate(gp_str = pt_sc2_pr$gp_str[match(lhs, pt_sc2_pr$label)]) |>
  filter(!is.na(gp_str)) |>
  arrange(desc(X2))
```

Free the second non-invariant threshold. 

```{r}
#| label: pr-scalar-partial-2
#| warning: false
fit_scalar_partial2_pr <- cfa(model_pr, data = pr, group = "site",
                              group.equal = c("loadings", "thresholds"),
                              group.partial = c("PR=~LonelySchool", "ClassNice | t3", "ClassHelp | t3"),
                              estimator = "WLSMV", ordered = pr_items)
fitMeasures(fit_scalar_partial2_pr, c("cfi", "tli", "rmsea", "srmr"))

fitMeasures(fit_metric_partial_pr, "cfi") - fitMeasures(fit_scalar_partial2_pr, "cfi")
```

ΔCFI is still > .01 — run the score test one more time to find the third non-invariant threshold.

```{r}
#| label: pr-scalar-score-3
#| warning: false
pt_sc3_pr <- parTable(fit_scalar_partial2_pr) |>
  filter(op == "|", group == 1) |>
  mutate(gp_str = paste0(lhs, "|", rhs))

lavTestScore(fit_scalar_partial2_pr)$uni |>
  mutate(gp_str = pt_sc3_pr$gp_str[match(lhs, pt_sc3_pr$label)]) |>
  filter(!is.na(gp_str)) |>
  arrange(desc(X2))
```

Free the third non-invariant threshold. This is the **final partial scalar model**.

```{r}
#| label: pr-scalar-partial-3
#| warning: false
fit_scalar_partial3_pr <- cfa(model_pr, data = pr, group = "site",
                              group.equal = c("loadings", "thresholds"),
                              group.partial = c("PR=~LonelySchool", "ClassNice | t3", "ClassHelp | t3", "ClassPlay | t3"),
                              estimator = "WLSMV", ordered = pr_items)
fitMeasures(fit_scalar_partial3_pr, c("cfi", "tli", "rmsea", "srmr"))

fitMeasures(fit_metric_partial_pr, "cfi") - fitMeasures(fit_scalar_partial3_pr, "cfi")
```

Print the full summary of the final model.

```{r}
#| label: pr-summary
summary(fit_scalar_partial3_pr)
```

Plot loadings (λ) and thresholds (τ) as a heatmap to compare estimates across sites.

```{r}
#| label: pr-params
#| fig-width: 8
#| fig-height: 3
pe_pr <- parameterEstimates(fit_scalar_partial3_pr)
gl_pr <- lavInspect(fit_scalar_partial3_pr, "group.label")

params_pr <- bind_rows(
  pe_pr |> filter(op == "=~") |>
    transmute(site  = factor(group, labels = gl_pr),
              item  = rhs, value = est, param = "lambda"),
  pe_pr |> filter(op == "|") |>
    transmute(site  = factor(group, labels = gl_pr),
              item  = lhs, value = est,
              param = paste0("tau", gsub("t", "", rhs)))
)

ggplot(params_pr, aes(param, item, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", value)), size = 3.5) +
  scale_fill_gradient2(low = "#2c7bb6", mid = "white",
                       high = "#d7191c", midpoint = 0, name = "Estimate") +
  scale_x_discrete(limits = c("lambda", "tau1", "tau2", "tau3"),
                   labels = c(lambda = expression(lambda),
                              tau1   = expression(tau[1]),
                              tau2   = expression(tau[2]),
                              tau3   = expression(tau[3]))) +
  facet_wrap(~ site, nrow = 1) +
  theme_classic() +
  labs(x = NULL, y = "Item")
```
