---
title: "SEM Tutorial"
format: pdf
---

This tutorial will give you a guided example of using CFA to test between different possible structures underlying a set of data. 
We use the Executive Function data from the LEVANTE Caregiver Survey - Child, aiming to determine how many factors best describe these data.

```{r}
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(pak, lavaan, tidySEM, tidyverse)
pak::pak("levante-framework/levante-r")
library(levante)
```

First, we download the data from the LEVANTE Redivis, and filter down to just the Executive Function (EF) construct.
```{r}
df_surveys <- get_surveys(data_source = "levante-data-pilots-hackathon",
                       version = "current")
df_surveys_ef <- df_surveys |> 
  filter(form_construct == "Executive Function")
```

Let's have a quick look at the questions in this construct.
```{r}
df_surveys_ef |> 
  select(form_subconstruct, variable, item_text) |> 
  distinct() |> 
  arrange(form_subconstruct, variable)
```

There are six subconstructs here, each of which has 5 or 6 items: 
1. Attentional focus, engagement, & persistence
2. Inhibitory control: Interference suppression
3. Inhibitory control: Response inhibition
4. Working memory
5. Cognitive flexibility
6. Planning and organisation

To conduct our CFA, we'll use the package `lavaan`, which expects one row per survey and one column per variable.
We will reshape our data into this wide format.
```{r}
df_surveys_ef_wide <- df_surveys_ef |> 
  select(survey_id, value, variable) |> 
  pivot_wider(names_from = "variable",
              values_from = "value")
```


We're going to do a CFA to determine whether these six subconstructs are the best structure of these variables. 
We'll compare the following three hypothesised models:
H1: Common factor model (all EF items load onto the same factor)
H2: 3-factor model (inhibition, working memory and attention, cognitive flexibility and planning)
H3: 6-factor model (each subconstruct is its own distinct factor)

Let's write out the model syntax for these hypothesised models.
`lavaan` uses the operator `=~` to indicate a factor relationship.

```{r}
h1_mod <- "
common_ef =~ ChildEFQ1 + ChildEFQ2 + ChildEFQ3 + ChildEFQ4 + ChildEFQ5 + ChildEFQ6 +        # attention
             ChildEFQ7 + ChildEFQ8 + ChildEFQ9 + ChildEFQ10 + ChildEFQ11 +                  # inhib - int supp
             ChildEFQ12 + ChildEFQ13 + ChildEFQ14 + ChildEFQ15 + ChildEFQ16 +               # inhib - resp inhib
             ChildEFQ17 + ChildEFQ18 + ChildEFQ19 + ChildEFQ20 + ChildEFQ21 + ChildEFQ22 +  # wm
             ChildEFQ23 + ChildEFQ24 + ChildEFQ25 + ChildEFQ26 + ChildEFQ27 + ChildEFQ28 +  # cog flex
             ChildEFQ29 + ChildEFQ30 + ChildEFQ31 + ChildEFQ32 + ChildEFQ33 + ChildEFQ34    # planning
"

h2_mod <- "
inhibition    =~ ChildEFQ7 + ChildEFQ8 + ChildEFQ9 + ChildEFQ10 + ChildEFQ11 +                  # inhib - int supp
                 ChildEFQ12 + ChildEFQ13 + ChildEFQ14 + ChildEFQ15 + ChildEFQ16                 # inhib - resp inhib
              
wm_attention  =~ ChildEFQ1 + ChildEFQ2 + ChildEFQ3 + ChildEFQ4 + ChildEFQ5 + ChildEFQ6 +        # attention
                 ChildEFQ17 + ChildEFQ18 + ChildEFQ19 + ChildEFQ20 + ChildEFQ21 + ChildEFQ22    # wm
                
flex_planning =~ ChildEFQ23 + ChildEFQ24 + ChildEFQ25 + ChildEFQ26 + ChildEFQ27 + ChildEFQ28 +  # cog flex
                 ChildEFQ29 + ChildEFQ30 + ChildEFQ31 + ChildEFQ32 + ChildEFQ33 + ChildEFQ34    # planning
"

h3_mod <- "
attention =~ ChildEFQ1 + ChildEFQ2 + ChildEFQ3 + ChildEFQ4 + ChildEFQ5 + ChildEFQ6          # attention
inhib_is  =~ ChildEFQ7 + ChildEFQ8 + ChildEFQ9 + ChildEFQ10 + ChildEFQ11                    # inhib - int supp
inhib_ri  =~ ChildEFQ12 + ChildEFQ13 + ChildEFQ14 + ChildEFQ15 + ChildEFQ16                 # inhib - resp inhib
wm        =~ ChildEFQ17 + ChildEFQ18 + ChildEFQ19 + ChildEFQ20 + ChildEFQ21 + ChildEFQ22    # wm
cog_flex  =~ ChildEFQ23 + ChildEFQ24 + ChildEFQ25 + ChildEFQ26 + ChildEFQ27 + ChildEFQ28    # cog flex
planning  =~ ChildEFQ29 + ChildEFQ30 + ChildEFQ31 + ChildEFQ32 + ChildEFQ33 + ChildEFQ34    # planning
"
```

Now let's fit these models, using the `lavaan::cfa()` function.
Note that there are missing data (not all caregivers filled in all items), so we ask `lavaan` to handle it using the `missing = "ML"` argument (case-wise maximum likelihood estimation).
```{r}
h1_cfa <- cfa(h1_mod, df_surveys_ef_wide, missing = "ML")
h2_cfa <- cfa(h2_mod, df_surveys_ef_wide, missing = "ML")
h3_cfa <- cfa(h3_mod, df_surveys_ef_wide, missing = "ML")
```

Now we can look at the fit measures of each model; we've chosen a few for illustration purposes but `lavaan` supports many more.
```{r}
# this function takes a while to compute
mod_fit <- map2(list(h1_cfa, h2_cfa, h3_cfa),
                c("H1", "H2", "H3"),
                \(f, n) fitmeasures(f, fit.measures = c("chisq", "cfi.robust", "bic", 
                                                        "rmsea", "rmsea.ci.lower", "rmsea.ci.upper")) |> 
                  as_tibble(rownames = "fit") |> 
                  mutate(model = n)) |> 
  list_rbind() |> 
  pivot_wider(names_from = "fit", values_from = "value")
mod_fit
```

Across all measures, the models with more factors appear to fit the data better (although the RMSEA values have overlapping confidence intervals).
Because these models are nested, we can also do a chi-square test to statistically compare them.
```{r}
anova(h1_cfa, h2_cfa, h3_cfa)
```
Recall that smaller chi-square is better; thus, in this case, our H3 model is significantly better than both H1 and H2.

Overall, these results confirm that a 6-factor structure of EF is better than a 1-factor or 3-factor structure!

We can also plot the factor loadings for our best model.
I've done a little bit of pre-cleaning to make the plot more legible, although the sheer number of items makes it quite challenging to read anyway.
Popping the diagram out in another window might help.
```{r}
graph_nodes <- get_nodes(h3_cfa) |> 
  mutate(label = str_remove(label, "ChildEF"))
graph_edges <- get_edges(h3_cfa) |> 
  mutate(show = !str_detect(label_results, "Variances."))
graph_sem(h3_cfa, nodes = graph_nodes, edges = graph_edges, text_size = 2)
```

We should also have a look at the actual factor scores, which we can do by printing the summary of the model.
```{r}
summary(h3_cfa)
```
Note that the factor loadings are all pretty close to 1 (anchor), and relatively even across items, suggesting that the different items all contribute similarly to each factor. Neat!
